Transfer learning for echelle spectroscopy

### 2021-11-05 paper/paper1/ms.tex
writing prose from outline, continued

We start with a high resolution pre-computed stellar model spectrum, $\mathsf{M}(\bm{\lambda})$.  For now, we require that the spectrum have sporadic regions of discernable continuum devoid of lines, and that the continuum varies smoothly in wavelength.  We suspect most stellar spectra meet these criteria, with the exception of the coolest M dwarfs and brown dwarfs.  The method is hypothetically capable of handling spectra without discernable continua, but we suspect some modifications to the preprocessing steps would be needed.  We allow the spectrum to have any number of spectral lines, so long as the continuum criterion is met.  We currently truncate the bandwidth of the precomputed synthetic spectra to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$, chosen to account for plausible radial velocity and rotational broadening of real stars.  A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

@@

\subsection{Initialization}
\begin{outline}
  \1 Initialize the model with preprocessing steps:
  \2 Low-pass filter to remove narrow spectral lines
  \2 (Optional) Place known Hydrogen, Sodium lines with large wings and non-standard line shapes
  \2 Define a prominence threshold and find locations of peaks above that threshold
  \2 Derive coarse properties about those peaks: widths and amplitudes
  \1 We now have a coarse model that needs to be fine-tuned with autodiff
\end{outline}

@@

% Table
\begin{deluxetable}{cc}
  \tablecaption{Notation used in this paper\label{table2}}
  \tablehead{
    \colhead{Symbol} & \nocolhead{Meaning}
  }
  \startdata
  $N_{pix}$ & Number of pixels in the precomputed spectrum \\
  $N_{lines}$ & Number of spectral lines \\
  $\sigma_G$ & Gaussian line profile scale and standard deviation \\
  $\gamma_L$ & Lorentzian line profile half width
  \enddata
\end{deluxetable}

>>>>

We start with a high resolution pre-computed stellar model spectrum, $\mathsf{M}(\bm{\lambda})$.  This spectrum should have sporadic regions of discernable continuum devoid of lines, and the continuum must vary smoothly in wavelength.  We suspect most stellar spectra meet these criteria, with the exception of the coolest M dwarfs and brown dwarfs.  The method can hypothetically handle spectra without discernable continua, but we suspect some modifications to the preprocessing steps would be needed.  We allow the spectrum to have any number of spectral lines.  We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$, chosen to account for plausible radial velocity and rotational broadening of real stars.  A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

@@

\subsection{Initialization}

We initialize the cloned model with a series of preprocessing steps.  We divide the entire spectrum by a black body of the same temperature as the model template.  This step alone is usually adequate to move on to the spectral line finding.

Occasionally a more advanced continuum flattening step may be needed to ensure that subsequent spectral line finding steps get applied uniformly.  Here, we apply a low-pass Savitzky-Golay filter\authorcomment1{citation XX} to identify the continuum regions, and then fit a low-order polynomial to this pre-filtered continuum spectrum.  We evaluate the low-order polynomial at all wavelength points, multiplied by the blackbody curve to assemble a ``perturbed black body'' that symbolizes the black body modulated by any broad-band continuum opacity that may reside in the model.  We define a threshold flux loss of $1\%$, below which any lines are masked, erode that mask by $N_{\mathrm{iter}}=13$ pixels, and repeat the polynomial fitting process again.  At this stage we fit a high order polynomial, with the polynomial order set through cross validation, typically $n_{\mathrm{poly}}\sim11$.  The ``perturbed black body'' continuum model $\mathsf{B}$ contains  $n_{\mathrm{poly}}+1$ free parameters.

In either continuum flattening case, we move on to find locations of spectral lines.
We apply a local minimum finding algorithm on the spectrum by defining a prominence threshold $\delta f\sim0.01$.
Testing demo with the real fixed static residual.

\begin{outline}
  \1 Define a prominence threshold and find locations of peaks above that threshold
  \2 Derive coarse properties about those peaks: widths and amplitudes
  \1 We now have a coarse model that needs to be fine-tuned with autodiff
\end{outline}

The low-pass filter should be set to capture the genuine spectral shape, without over-fitting broad line-wings such as those in deep Hydrogen and sodium lines.  We have experimentally found it helpful to hand-tune models of exceptionally broad and non-standard line-wings with with an optional pre-processing step.

@@

% Table
\begin{deluxetable}{cc}
  \tablecaption{Notation used in this paper\label{table2}}
  \tablehead{
    \colhead{Symbol} & \nocolhead{Meaning}
  }
  \startdata
  $N_{pix}$ & Number of pixels in the precomputed spectrum \\
  $N_{lines}$ & Number of spectral lines \\
  $\sigma_G$ & Gaussian line profile scale and standard deviation \\
  $\gamma_L$ & Lorentzian line profile half width\\
  $\pm \Delta \lambda_{\mathrm{buffer}}$ & Buffer exceeding the red and blue limits of data spectrum
  \enddata
\end{deluxetable}
